Predicting instabilities of a tuneable ring laser with an iterative map model

### 2021-06-02 Papers/Journal/Tuneable_Lasers.tex
Removed instances of modulation instability

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions leading to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component in the laser cavity, we obtain an algebraic map for each component. Iterating through the maps gives the total effect of one round trip. By neglecting the nonlinearity, we find a linearly chirped Gaussian to be the analytic fixed point solution, which we analyze asymptotically. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity as the pulse evolves over hundreds of round trips of the cavity. In the nonlinear case, we find the chirp saturates, and the Fourier transform of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region and the unstable region where modulation instability degrades the pulse into an unsustainable state.
\end{abstract}

@@

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Common cavity construction of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The laser travels clockwise around the loop, with most of the pulse exiting the cavity through the optical coupler. CFBG, chirped fibre Bragg grating; OC, optical circulator; WDM, wavelength division multiplexer}
	\label{fig:cavity}
\end{figure}

Several effects arise within the cavity due to the interplay of dispersion, modulation, and nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for us are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to redshift, and the trailing edge to blueshift (the effect reverses in the anomalous dispersion regime) to the point a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking arises from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into multiple, much shorter, less intense pulses and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability can occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity, wave breaking and modulation instability can become parasitic, leading to an unstable and unsustainable pulse. Understanding the rich landscape of the parameter space is important for determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

@@

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. A linear iterative method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the interactions of dispersion, modulation, and nonlinearity---hindering their models' ability to exhibit modulation instability.

In this paper, we build an iterative map model for the evolution of a laser pulse as the pulse travels through the cavity. In Section~\ref{sec:model}, we derive our own iterative map from \eqref{eq:nlse}, we non-dimensionalize the model and discover four non-dimensional parameters that govern the dynamics of the system. We then examine the results of our model in Section~\ref{sec:results}---first solving the model analytically by assuming the absence of the nonlinearity, then numerically solving the full nonlinear model. A sharp boundary is identified in the parameter space separating the region of stability, and the region where modulation instability degrades the pulse. Finally, our concluding thoughts and ideas for possible future work are given in Section~\ref{sec:conclusion}.

@@

The geometry of a particular laser will dictate the order of the process maps. However, we assume the construction of the laser is as shown in Fig.~\ref{fig:cavity}. The effect of the nonlinearity will be maximal after the pulse passes through the gain fibre---the pulse is most energetic after being amplified by the gain. Thus, we take the nonlinearity component to follow the gain. Similarly, we want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. All that remains is the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our starting point since the loss corresponds to the optical coupler, and thus, allows us to more easily compare our results with experiments. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reach equilibrium, or the pulse succumbs to modulation instability. The phase change is uninteresting to us, as the phase cannot easily be measured experimentally. An illustration of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

In Fig.~\ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} after reaching equilibrium for the nominal parameter values given in \eqref{eq:ndparam}, and two additional $b$ values. In the nonlinear case, we find the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transforms. Notice, in comparison to the linear case, the nonlinearity introduces higher frequency modes, giving a bi-modal distribution. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| \gtrsim 1$, the chirp begins to saturate---consistent with the experimental results~\cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, increases, the chirp increases more sharply across the pulse, and saturates at a larger value. Also, the two peaks in the Fourier transform become more pronounced and further apart, and the Fourier transform of the pulse becomes more rectangular. However, after a critical point, SPM plays a more substantial role, leading to modulation instability.

Fig.~\ref{fig:breakevo} highlights the phenomenon of modulation instability. By decreasing the dispersion compared to Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity the SPM compounds and becomes detrimental to the pulse. In turn, modulation instability is induced, and degrades the pulse, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) forms, which is able to grow because of the gain medium. The central lobe then allows the pulse to come to equilibrium quickly. Furthermore, since the intensity of the initial pulse is small relative to the saturation energy, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. Evidently, the initial shape of the pulse is much less important than the initial energy of the pulse, or the strength of the nonlinearity.

@@

We now turn our attention to a more quantitative analysis, and characterize the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular, since $a$ and $h$ affect only the amplitude of the pulse, whereas the interactions of dispersion and nonlinearity give rise to the instabilities. We examine the relative change in the pulse's envelope between consecutive round trips of the cavity to identify stability. We compute the error as
\begin{align}
	\Delta = \frac{\| |\mathcal{L}^n(A_0)| - |\mathcal{L}^{n-1}(A_0)| \|_2}{\| \mathcal{L}^{n-1}(A_0) \|_2},
	\label{eq:error}
\end{align}
where
\begin{align}
	\| f \|_2^2 \coloneqq \int_{-\infty}^\infty |f|^2 \, \df T,
\end{align}
and we choose $n$ to be sufficiently large to guarantee the pulse reaches equilibrium or deteriorates because of modulation instability---we use $n = 500$ in our experiments. We take the modulus of the pulse since we are interested in the evolution of only the envelope because of the auto-correlation methods used for experimental measurements. We plot the error in Fig.~\ref{fig:error}. Two distinct regions split the parameter space, divided by an incredibly sharp, and complicated boundary. In the upper-left region, the error is $\bigO{1}$; here, the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable to within machine precision. The increase in dispersion, coupled with modulation, allows the laser to balance the nonlinear effects.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Given our ordering of the components, we compute the energy of the pulse immediately before passing through the optical coupler. Thus, the energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is small. The pulse is not able to foster prominent modes in this region, and is unsustainable, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find the energy decays smoothly as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, the tails of the pulse increase in length due to the dispersion, and more energy is lost due to modulation. As $b$ increases, our choice of modulation function helps suppress the additional frequencies generated by SPM, ensuring the pulse remains band-limited. We find a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse. Moreover, the maximum energy, which occurs in the limit where both $s$ and $b$ go to zero, is 2.5487. With our choice of parameters, \eqref{eq:maxenergy} predicts an equilibrium energy of 2.5535 at the origin, confirming the usefulness of the approximation. Moreover, we show in Fig. \ref{fig:energytime} the energy variation of four pulses during the 500 round trips. In particular, we plot two pulses in the stable region (purple and blue) and two pulses in the unstable region (green and red). Each of the pulses, again, has an initial energy of $E_0 = 0.1$, and after about a dozen round trips the energy is approximately 1.25 in the case of the $s = 0.35$, $b = 6.0$, and approximately 2 in the other three cases. The two examples in the stable region quickly reach equilibrium, and their energy remains constant for the remaining round trips. However, for the examples in the unstable region the pulses lose a considerable amount of their energy and the pulse begins to degrade. The energy of the pulses varies by up to 25\% between consecutive round trips. A pulse in the unstable region undergoes large transformations with each round trip leading to the error we saw in Fig. \ref{fig:error}, as well as the energy variation in Fig. \ref{fig:energytime}.

@@

Analyzing the nonlinear model analytically using an asymptotic expansion for small values of the nonlinearity parameter, $b$, is of interest. Doing so will provide insight to how the nonlinearity impacts the linear solution to better understand the manifestation of wave breaking and modulation instability. Furthermore, a more detailed analysis of crossing the boundary between stability and instability may provide valuable insight for mechanisms that trigger the instability.

>>>>

\begin{abstract}
	Simple mathematical models have been unable to predict the conditions leading to instabilities in a tuneable ring laser. Here, we propose a nonlinear iterative map model for tuneable ring lasers. Solving a reduced nonlinear Schr\"odinger equation for each component in the laser cavity, we obtain an algebraic map for each component. Iterating through the maps gives the total effect of one round trip. By neglecting the nonlinearity, we find a linearly chirped Gaussian to be the analytic fixed point solution, which we analyze asymptotically. We then numerically solve the full nonlinear model, allowing us to probe the underlying interplay of dispersion, modulation, and nonlinearity as the pulse evolves over hundreds of round trips of the cavity. In the nonlinear case, we find the chirp saturates, and the Fourier transform of the pulse becomes more rectangular in shape. Finally, for a nominal plane in the parameter space, we uncover a rich, sharp boundary separating the stable region, and the unstable region where instabilities degrade the pulse into an unsustainable state.
\end{abstract}

@@

\begin{figure}[tbp]
	\centering
	\input{Cavity}
	\caption{Common cavity construction of a fibre ring laser~\cite{burgoyne2014, chung2017, lapre2019, shao2019}. The laser travels clockwise around the loop, with most of the pulse exiting the cavity through the optical coupler. CFBG, chirped fibre Bragg grating; OC, optical circulator; WDM, wavelength division multiplexer.}
	\label{fig:cavity}
\end{figure}

Several effects arise within the cavity due to the interplay of dispersion, modulation, and nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. One effect, wave breaking, causes the leading edge of a pulse to redshift, and the trailing edge to blueshift (the effect reverses in the anomalous dispersion regime) to the point a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking arises from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Outside of optics, wave breaking can also occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity, combination of dispersion, modulation, and nonlinearity can also lead to instabilities that degrade the pulse. Wave breaking and instabilities can become parasitic, leading to an unstable and unsustainable pulse. Understanding the rich landscape of the parameter space is important for determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

@@

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. A linear iterative method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the interactions of dispersion, modulation, and nonlinearity---hindering their models' ability to exhibit instabilities.

In this paper, we build an iterative map model for the evolution of a laser pulse as the pulse travels through the cavity. In Section~\ref{sec:model}, we derive our own iterative map from \eqref{eq:nlse}, we non-dimensionalize the model and discover four non-dimensional parameters that govern the dynamics of the system. We then examine the results of our model in Section~\ref{sec:results}---first solving the model analytically by assuming the absence of the nonlinearity, then numerically solving the full nonlinear model. A sharp boundary is identified in the parameter space separating the stable region from the unstable region. Finally, our concluding thoughts and ideas for possible future work are given in Section~\ref{sec:conclusion}.

@@

The geometry of a particular laser will dictate the order of the process maps. However, we assume the construction of the laser is as shown in Fig.~\ref{fig:cavity}. The effect of the nonlinearity will be maximal after the pulse passes through the gain fibre---the pulse is most energetic after being amplified by the gain. Thus, we take the nonlinearity component to follow the gain. Similarly, we want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. All that remains is the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our starting point since the loss corresponds to the optical coupler, and thus, allows us to more easily compare our results with experiments. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reach equilibrium, or the pulse becomes unstable. The phase change is uninteresting to us, as the phase cannot easily be measured experimentally. An illustration of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

In Fig.~\ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} after reaching equilibrium for the nominal parameter values given in \eqref{eq:ndparam}, and two additional $b$ values. In the nonlinear case, we find the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transforms. Notice, in comparison to the linear case, the nonlinearity introduces higher frequency modes, giving a bi-modal distribution. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| \gtrsim 1$, the chirp begins to saturate---consistent with the experimental results~\cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, increases, the chirp increases more sharply across the pulse, and saturates at a larger value. Also, the two peaks in the Fourier transform become more pronounced and further apart, and the Fourier transform of the pulse becomes more rectangular. However, after a critical point, SPM plays a more substantial role, leading to instability.

Fig.~\ref{fig:breakevo} highlights the instability. By decreasing the dispersion compared to Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity the SPM compounds and becomes detrimental to the pulse. In turn, the pulse starts to become unstable and further degrades, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) forms, which is able to grow because of the gain medium. The central lobe then allows the pulse to come to equilibrium quickly. Furthermore, since the intensity of the initial pulse is small relative to the saturation energy, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. Evidently, the initial shape of the pulse is much less important than the initial energy of the pulse, or the strength of the nonlinearity.

@@

We now turn our attention to a more quantitative analysis, and characterize the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular, since $a$ and $h$ affect only the amplitude of the pulse, whereas the interactions of dispersion and nonlinearity give rise to the instabilities. We examine the relative change in the pulse's envelope between consecutive round trips of the cavity to identify stability. We compute the error as
\begin{align}
	\Delta = \frac{\| |\mathcal{L}^n(A_0)| - |\mathcal{L}^{n-1}(A_0)| \|_2}{\| \mathcal{L}^{n-1}(A_0) \|_2},
	\label{eq:error}
\end{align}
where
\begin{align}
	\| f \|_2^2 \coloneqq \int_{-\infty}^\infty |f|^2 \, \df T,
\end{align}
and we choose $n$ to be sufficiently large to guarantee the pulse reaches equilibrium or deteriorates because of the instability---we use $n = 500$ in our experiments. We take the modulus of the pulse since we are interested in the evolution of only the envelope because of the auto-correlation methods used for experimental measurements. We plot the error in Fig.~\ref{fig:error}. Two distinct regions split the parameter space, divided by an incredibly sharp, and complicated boundary. In the upper-left region, the error is $\bigO{1}$; here, the nonlinearity is too strong, and the instability is induced by SPM. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable to within machine precision. The increase in dispersion, coupled with modulation, allows the laser to balance the nonlinear effects.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Given our ordering of the components, we compute the energy of the pulse immediately before passing through the optical coupler. Thus, the energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is small. The pulse is not able to foster prominent modes in this region, and is unsustainable, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find the energy decays smoothly as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, the tails of the pulse increase in length due to the dispersion, and more energy is lost due to modulation. As $b$ increases, our choice of modulation function helps suppress the additional frequencies generated by SPM, ensuring the pulse remains band-limited. We find a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise the pulse will succumb to the instability. Moreover, the maximum energy, which occurs in the limit where both $s$ and $b$ go to zero, is 2.5487. With our choice of parameters, \eqref{eq:maxenergy} predicts an equilibrium energy of 2.5535 at the origin, confirming the usefulness of the approximation. Moreover, we show in Fig. \ref{fig:energytime} the energy variation of four pulses during the 500 round trips. In particular, we plot two pulses in the stable region (purple and blue) and two pulses in the unstable region (green and red). Each of the pulses, again, has an initial energy of $E_0 = 0.1$, and after about a dozen round trips the energy is approximately 1.25 in the case of the $s = 0.35$, $b = 6.0$, and approximately 2 in the other three cases. The two examples in the stable region quickly reach equilibrium, and their energy remains constant for the remaining round trips. However, for the examples in the unstable region the pulses lose a considerable amount of their energy and the pulse begins to degrade. The energy of the pulses varies by up to 25\% between consecutive round trips. A pulse in the unstable region undergoes large transformations with each round trip leading to the error we saw in Fig. \ref{fig:error}, as well as the energy variation in Fig. \ref{fig:energytime}.

@@

Analyzing the nonlinear model analytically using an asymptotic expansion for small values of the nonlinearity parameter, $b$, is of interest. Doing so will provide insight to how the nonlinearity impacts the linear solution to better understand the manifestation of wave breaking and the instability. Furthermore, a more detailed analysis of crossing the boundary between stability and instability may provide valuable insight for mechanisms that trigger the instability.
